---
title: "Lab12: during lab activities"
author: "François Birgand"
date: '2023-03-30'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this lab, you are welcome to use all code source I have provided to you before and during lab as appropriate.

# Exercise 1 

Plot the yearly hydrograph for the Cuyahoga River at Independence, Ohio. For this, copy/paste the correct code from a code cell from `BAE204-QClab.Rmd` and paste into a *code chunk* below. I have added a blank code chunk below you can use throughout. Make sure you adjust the x-axis label such that it reflects the years. Make sure you modify the title as well such that it reads *Flow rates at Cuyahoga River at Independence, OH*. 


```{r echo=FALSE, eval=TRUE}
# this is a blank code chunk. You can have/or not have the code appear in the knitted document by switching echo=TRUE or echo=FALSE within the {} of the chunk first line
# Remove these comments when you copy/paste your code.

data<-read.csv(file = "Lin_1h_cuyahoga_all_2008-2009_SI_ref.csv", header = TRUE)

#data<-read.csv(file="sample_1hr_QC_data.csv",header = TRUE) #Reads file into table format
  
  WSarea<-24.2 #Area of watershed in km2
  WS<-"Flow rates at Cuyahoga River at Independence, OH"
  #names(data)=c("datetime","Q","C")   # renames the columns in simpler names
  data<-as.data.frame(data)
  data$datetime<-as.POSIXct(strptime(data$datetime, "%Y-%m-%d %H:%M:%S")) # transforms characters into date values understood by R
  D<-data$datetime
  Q<-data$Q   #Defines Q as the flow value (m3/s)
  
  N=nrow(data)   #Sets N to the value equal to the number of total rows in the table
  
  # definition of the x and y axes limits
  
  startdate<-D[1]
  enddate<-D[N]
  xlim = as.POSIXct(c(startdate,enddate))  # this renders the first and last date understandable for plotting purposes
  ylimQ = c(0,max(Q))           # ylim for flow
      
  ScaleF = 1.2                  # scaling factor for size of fonts and other things
      
  y1lab<-expression("Flow rate (" * m^3 * "/s)")  # defines the label for flow
  
  par(mar=c(4.5,4.5,4,4.5))     # defines the sizes, in number of lines, for the margins (bottom, left, top, right)
  
      ltyp=c(1,2)
      
      plot(D,Q,col="blue",type="l",cex=0.1,yaxt="n",     
           lty=ltyp[1],xaxt="n",xlab="",ylab="",xlim=xlim,ylim=ylimQ)
      # we are taking all the default addition of axis tick marks and numbers out by using xaxt and yaxt = "n"
      # and setting the axis labels at nothing using xlab = "" and ylab = ""
      abline(h=0)
      abline(v=seq(startdate, enddate, by="week"),col=("grey"))
      axis.POSIXct(1, at=seq(startdate, enddate, by="month"), format="%m/%d",cex.axis=ScaleF)
      # this tells R that we want the X axis ticks and values to be displayed as dates, be added on a monthly basis,
      # using the month/day format
      axis(2,cex.axis=ScaleF)
      # this tells R that the first Y axis ticks can be displayed  (that function was repressed earlier by 'yaxt="n" ')
      par(new=TRUE)
      # this tells R that a new plot has already been opened, in other words you are telling R to keep adding things
      # on the existing plot
      
      mtext("Dates in 1997-1998",side=1,line=3,cex=ScaleF) # add in the margin the defined labels and title
      mtext(y1lab,side=2,line=3,cex=ScaleF)
      mtext(WS,side=3,line=1.5,cex=ScaleF)
      
      legend("topleft",c("Flow"),lty = c(1), col = c("blue"))
   
  

```

# Exercise 2

Calculate the lowest, highest, mean, median, as well as the $10^{th}$ and $90^{th}$ percentile of the flow rates. Make sure you use 3 significant digits. To report a number `N` with 4 significant digits, the syntax would be `signif(N,4)`.

In `R`, the code to calculate the mean of a vector `X` is `mean(X)`.
To calculate the, e.g. 24% percentile of a variable X, the syntax is `quantile(X,0.24)`

The mean flow rate was `r signif(mean(Q),3)` m^3^/s

An example for the code in Colab would be: 

```{r echo=FALSE, eval=FALSE}
meanQ = mean(Q)
paste0("The mean flow was ", signif(meanQ,digits=3)," m^3/s")
```

# Exercise 3 

Calculate the arithmetic average and medium concentrations for nitrate, TP, SRP, and TSS over the entire year of record. SRP stands for Soluble Reactive Phosphorus, and it essentially corresponds to phosphate. You will see that for phosphate, there are some negative values floating around that are just erroneous. You need to remove them, replace them with `NA` values and calculate the mean omitting the `NA` values in your calculation. For that, use the syntax below. Assuming that PO4 would be the name of the SRP vector, you would use:

```{r echo=FALSE, eval=FALSE}
PO4[PO4<=0]=NA # this tells R that all values lesser or equal to 0 are to be replaced by NA
mean(na.omit(PO4)) # this tells R to calculate the mean while omitting the NA values in the calculation. 
#If you do not specify this, R is going to give an error message.
```

Compare your values to the reference values for agricultural watersheds, which are, respectively, 5 mg N/L, 0.3 mg P/L, 0.03 mg P/L, and 20 mg TSS/L. 

Report your results (in R Markdown) using this syntax and replace the `___` by embedded code

- The mean concentration for nitrate was ___ mg N/L
- The median concentration for nitrate was ___ mg N/L.

And so on. Compare mean versus median and discuss the difference. And there almost the same, or not. If they were really different, think about what would is say about the occurrence of high values. Compare them to the reference values we have seen in class. Similarly discuss your findings for the other parameters.

# Exercise 4

Plot the hydrograph and nitrate, TP, SRP, and TSS chemographs for the Cuyahoga River at Independence, Ohio, over the first two weeks of March 2009. You should make 4 plots for that. To 'zoom in' to these two weeks, in the code we used in lab and you probably pasted for exercise 1, the xlim are defined by 'startdate' and 'enddate'. Modify them accordingly.

Using text, report whether the chemographs exhibit a concentration or a dilution effect.



# Exercise 5

Plot the cumulative load of nitrate, TP, SRP, and TSS as a function of cumulative flow volume for the full year for the Cuyahoga River at Independence, Ohio. You should make 4 plots for that, and add a straight line between the first and last point. Make sure you modify the axis title to reflect what you are plotting with the correct units: kg on Y-axis, m^3^ on X-axis. For SRP, you should replace all negative values by 0.005 mg P/L (using the same procedure detailed above), otherwise, you will not be able to calculate the cumulative loads.

Calculate the annual flow weighted average concentrations (FWAC) for nitrate, TP, SRP, and TSS, and compare them to the arithmetic average. Bonus +2: what higher/lower FWAC suggests on concentration/dilution effects.



# Exercise 6

Plot the flow volume duration curve over the entire period of record.
Report the total flow volume that occurred in 2%, 5%, and 10% of the time, corresponding to the highest flow.



# Exercise 7 

Plot the cumulative load duration curves for nitrate, phosphate, TP, and TSS over the entire period of record all on the same graph. Make sure you change colors.
Report the total load that occurred in 2%, 5%, and 10% of the time, corresponding to the highest loads for each of the parameters .

Compare these values to the ones you obtained for flow, and discuss whether or not that makes sense.



